Research log

### 2025-10-09 docs/paper.Rmd
model fits merged into one file, saving intermediate data automatically, and beautifying some parts of the paper

---
title             : "Pretending not to know reveals a powerful capacity for self-simulation"
shorttitle        : "Pretending not to know"

author: 
  - name          : "Matan Mazor"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "All Souls College, High Street, Oxford OX1 4AL"
    email         : "matan.mazor@all-souls.ox.ac.uk"
  #   role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
  #     - Conceptualization
  #     - Writing - Original Draft Preparation
  #     - Writing - Review & Editing

@@

  - name          : "Ian Phillips"
    affiliation   : "2"
    # role:
    #   - Writing - Review & Editing

@@

abstract: |
  What we do depends on what we know. But sometimes we try to decouple our behavior from our knowledge so as to appear not to know what we really do know. Such pretense behavior requires understanding how we would behave with different knowledge, and so provides a window into counterfactual self-simulation. However, little research has characterized and evaluated pretense relative to non-pretense behavior. Here, in two large-scale, pre-registered experiments, we tested people’s ability to conceal knowledge in the context of information-seeking games (Battleship and Hangman). Participants played normal games (in which they began with no knowledge of the ships’ locations or the word’s identity), as well as ‘pretend’ games in which they were shown the game's full solution but were instructed to play *as though* they were ignorant. We find a remarkable capacity for pretense, with pretenders successfully reproducing broad patterns and subtle features of real players’ decision-making and timing — so much so that 'judge' participants failed to distinguish real players from pretenders. Nonetheless, pretense was characterized by pervasive differences, undetected by judges but revealed by follow-up analyses: pretenders played less rationally than real players, and had an increased tendency to rely on heuristics and rules. Our findings thus reveal a powerful but imperfect ability to simulate decisions and behavior based on a counterfactual belief state, and open up a new approach to studying self-simulation.

  <!-- https://tinyurl.com/ybremelq -->

keywords          : "pretense; metacognition; theory of mind"
wordcount         : "5986"

@@

classoption       : "man"
output            : papaja::apa6_word
bibliography: references.bib
---

```{r analysis-preferences, warning=F, message=F}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed, warning=F, message=F)

@@

# Load workspace (after running preregisteredMethodsAndResults.Rmd, 
# exploratoryResults.Rmd, and all the scripts that are linked to from these 
# documents). 
load('../.RData')

r_refs("r-references.bib")

@@

@@

<!-- The ability to intentionally deceive others relies on a capacity to reason about mental states [@frith2005theory]. This is evident in a similar developmental trajectory for the acquisition of theory of mind and the ability to deceive and detect deception [@wimmer1983beliefs; @shultz1981development; @sodian1991early], and a similar distribution of deception and theory of mind in the animal kingdom [e.g., @emery2004mentality; @hall2017cooperation]. This link makes conceptual sense: to deceive others, one needs to understand that others can have different knowledge and beliefs than one's own. -->

Pretense relies on an ability to simulate and mimic one's own behavior under a counterfactual belief state. For example, in order to successfully deceive your friends into thinking that you were surprised by the birthday party they threw for you, it is not sufficient that you are able to reason about their mental states ("I know that they are planning a surprise party, but they don't know that I know that...") --- you also need to convincingly simulate and mimic your hypothetical behavior had you not known about the party ("Where would I look first had I not known? What would I say? How long would it take me to recover from the surprise?"). This is not a trivial challenge: previous research on "hindsight biases" suggests that knowledge about the actual state of the world can interfere with our ability to correctly judge what we would have believed [@fischhoff1975; @fischhoff1977; @wood1978; @roese2012] or perceived [@harley2004; @bernstein2007; @bernstein2012] without this knowledge. Such biases remain potent even when instructing participants to overcome them [@harley2004; @pohl1996]. Moreover, even if pretenders can correctly determine what they would have believed, they must further accurately simulate how they would think and behave in this different belief state.

@@

<!-- to what extent humans are capable of generating complex behavior that is aligned with a counterfactual belief state, and specifically with a counterfactual state of ignorance. -->

To this end, we examine pretense in a game setting. Using an online version of the games Battleship and Hangman (in which players seek to uncover the locations of enemy ships or the identity of a word), participants played a 'non-pretend' (normal) version of the game, as well as a 'pretend' version where they were given complete information about the hidden ships / target word but were instructed to behave as if they didn't have this information. Overall, we find a highly impressive capacity for pretending not to know, with pretenders mirroring broad patterns and subtle features of real players' decisions and decision times. However, we also find that pretense was characterized by over-acting, stereotypical behavior, and suboptimal incorporation of new information, although these were undetected by human observers.

@@

In two experiments, players played online versions of two information-seeking games: Battleship (N=`r E2.df$subj_id%>%unique()%>%length()` players) and Hangman (N=`r E4.df$subj_id%>%unique()%>%length()` players). These games traditionally start in a state of ignorance, with a player's goal being to reveal an unknown world state (ship locations in Battleship, a hidden word in Hangman) in as few steps (cell or letter selections) as possible. Critically, in addition to playing five standard games, players in our experiments also completed five 'pretend' games in which the solution to the game was known to them from the start, but where their task was to behave as if it was not -- i.e. to play as though they did not have this information (see Fig. \@ref(fig:design)). We measured participants' capacity to simulate a counterfactual state of ignorance by comparing patterns of decisions and decision times in pretend and non-pretend games. Our full pre-registered results are available [online](https://matanmazor.github.io/ignorance/docs/exp.-1-battleship.html) together with the report-generating code. Readers are invited to try [demos of the experiments](https://matanmazor.github.io/ignorance/experiments/demos/pretend).

@@

As a first measure of pretense quality, we compared the total number of guesses in pretend and non-pretend games. Among Battleship players, the number of cell selections was similar in pretend (mean = `r E2.median_clicks%>%pull(pretend)%>%mean()%>%printnum()`) and non-pretend games (mean = `r E2.median_clicks%>%pull(nonpretend)%>%mean()%>%printnum()`; $p=$ `r apa_print(E2.median_clicks$diff%>%t.test())$table$p.value`; Fig. \@ref(fig:decisions)A). `r E2.lucky_pretenders%>%length` pretenders who immediately discovered all ships without making errors were excluded from all further analyses, in accordance with our pre-registered plan. With these subjects excluded, the number of cell selections remained very similar in pretend (mean = `r  E2.median_clicks_filtered%>%pull(pretend)%>%mean()%>%printnum()`) and non-pretend games (mean = `r E2.median_clicks_filtered%>%pull(nonpretend)%>%mean()%>%printnum()`; $p=$`r apa_print(E2.median_clicks%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$table$p.value`; Fig. \@ref(fig:decisions)A)). In Hangman, pretenders tended to make about one additional letter guess on average than did non-pretenders, controlling for word length (pretend: `r E4.median_misses%>%pull(pretend)%>%mean()%>%printnum()` misses; non-pretend: `r E4.median_misses%>%pull(nonpretend)%>%mean()%>%printnum()` misses; $p$ `r apa_print(E4.median_misses%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$table$p.value`; Fig. \@ref(fig:decisions)B). Despite an overall bias in the number of guesses, pretend Hangman games showed an impressive, item-specific alignment: pretenders were successful in making more letter guesses when attempting to reveal words that would have been harder to guess had they been playing for real (`r cor.test(E4.misses_per_word$pretend, E4.misses_per_word$nonpretend,method='spearman')%>%apa_print()%>%'$'(estimate)`; Fig. \@ref(fig:decisions)B). This strong correlation provides evidence for a human capacity to act in accordance with a counterfactual knowledge state.

(ref:decisions-caption) Battleship and Hangman guesses in pretend and non-pretend games. A: median number of misses in Battleship and Hangman games, in non-pretend (green) and pretend (purple) games. For reference, the expected number of misses is indicated by a reference line for a fully random agent, and for a "greedy" agent that maximizes the probability of a hit in each step. B: The median number of misses in Hangman for pretend and non-pretend games, as a function of the target word. C: Spatial guess distributions for pretend and non-pretend half-games (where players continued the game from a half-finished state) alongside their corresponding hit probability maps. D: Cell and letter selections were ranked according to their relative hit probability. Plotting the median rank per subject in pretend and non-pretend games, with reference lines for the expected rank probability for a random agent, and for a "greedy" agent that maximizes the probability of a hit in each step. E: same as panel D but discarding all guesses that resulted in a hit.

@@

num_misses_df <- E2.click_df %>%
  dplyr::select(subj_id,test_part, num_clicks) %>%
  mutate(subj_id=as.character(subj_id)) %>%
  # rbind(E2.optimal_click_df) %>%
  # rbind(E2.random_click_df) %>%
  mutate(test_part = factor(test_part, levels=c(
    'nonpretend','pretend'), labels = c(
    'non-pretend','pretend'))) %>%
  group_by(subj_id,test_part) %>%
  summarise(num_misses=median(num_clicks)-7) %>%
  mutate(game='Battleship') %>%
  rbind(
    E4.click_df %>%
  dplyr::select(subj_id,test_part, num_misses) %>%
  mutate(subj_id=as.character(subj_id)) %>%
  # rbind(E2.optimal_click_df) %>%
  # rbind(E2.random_click_df) %>%
  mutate(test_part = factor(test_part, levels=c(
    'nonpretend','pretend'), labels = c(
    'non-pretend','pretend'))) %>%
  group_by(subj_id,test_part) %>%
  summarise(num_misses=median(num_misses)) %>%
    mutate(game='Hangman')
  )
load('../analysis/simulated_hangman_games.RData')

@@

p<-  E2.click_df %>%
  dplyr::select(subj_id,test_part, num_clicks) %>%
  mutate(subj_id=as.character(subj_id)) %>%
  # rbind(E2.optimal_click_df) %>%
  # rbind(E2.random_click_df) %>%
  mutate(test_part = factor(test_part, levels=c(
    'nonpretend','pretend'), labels = c(
    'non-pretend','pretend'))) %>%
  group_by(subj_id,test_part) %>%
  summarise(num_misses=median(num_clicks)-7) %>%
  ggplot(aes(x=num_misses,fill=test_part, color=test_part))+
  scale_fill_manual(values=c(colors$np,colors$p, colors$random, colors$greedy)) +
  scale_x_continuous(limits=c(-1,19), breaks=seq(0,18)) +
  labs(x='median number of misses', y='number of players')+
  theme(legend.position='none') +
  theme_classic() +
  geom_vline(xintercept = E2.greedy_mean_misses, color='black', linetype=2) +
  annotate("text", x = E2.greedy_mean_misses, y = 75, label = "Greedy", angle = 90, vjust = -1, size = 5)+
  geom_vline(xintercept = E2.random_mean_misses, color='black', linetype=2) +
  annotate("text", x = E2.random_mean_misses, y = 75, label = "Random", angle = 90, vjust = -1, size = 5)+
  geom_bar(alpha=0.6, position = 'identity', color='black', size=0.5) +
  theme(legend.pos='na')

@@

Having established an alignment in the total number of guesses, we next turned to the content of pretend and non-pretend guesses. In order to directly compare pretend and non-pretend guesses for the same board state, Battleship players completed two half-games in which they were instructed to continue the game from a half-completed state. Hundreds of cell selections for the same board state revealed a strong correlation between the spatial distributions of pretend and non-pretend guesses (board A: $r=$ `r apa_print(cor_A_pretend_nonpretend)$table$estimate`, $p=$ `r apa_print(cor_A_pretend_nonpretend)$table$p.value`; board B: $r=$ `r apa_print(cor_B_pretend_nonpretend)$table$estimate`, $p$ `r apa_print(cor_B_pretend_nonpretend)$table$p.value`; Fig. \@ref(fig:decisions)C), confirming that pretenders were sensitive not only to the number of guesses they would have made had they been playing for real, but also to their content.

To further examine the decisional processes behind this strong alignment, we compared the degree to which pretend and non-pretend guesses made sense within the context of the game. When playing Battleship and Hangman, it makes sense to guess cells or letters for which the probability of hitting a ship or revealing a letter is high [this "greedy" behavior is not strictly optimal, but approximates optimal behavior in most cases, @audinot2014optimal, Section 3.3]. In the non-pretend versions of both games, guesses were more rational according to this measure than expected by chance (Battleship: `r E2.mean_P_click_rank%>%spread(test_part,p_click_rank)%>%mutate(diff=-nonpretend+random)%>%pull(diff)%>%t.test()%>%apa_print()%>%'$'(statistic)`, Hangman: `r E4.mean_P_click_rank%>%spread(test_part,p_click_rank)%>%mutate(diff=-nonpretend+random)%>%pull(diff)%>%t.test()%>%apa_print()%>%'$'(statistic)`). Despite being less rational than non-pretend guesses, pretend guesses were also more rational than expected by chance (Battleship: `r E2.mean_P_click_rank%>%spread(test_part,p_click_rank)%>%mutate(diff=-pretend+random)%>%pull(diff)%>%t.test()%>%apa_print()%>%'$'(statistic)`, Hangman: `r t.test(-E4.mean_P_click_rank_wide%>%pull(np_vs_r))%>%apa_print%>%'$'(statistic)`; Fig. \@ref(fig:decisions)D). Critically, the same was true when restricting the analysis to unsuccessful guesses (Battleship: `r E2.mean_P_click_rank_misses_only%>%spread(test_part,p_click_rank)%>%mutate(diff=-pretend+random)%>%pull(diff)%>%t.test()%>%apa_print()%>%'$'(statistic)`, Hangman: `r t.test(-E4.mean_P_click_rank_misses_wide%>%pull(p_vs_r))%>%apa_print%>%'$'(statistic)`; Fig. \@ref(fig:decisions)E): that is, even when incorrectly guessing a ship's location or a letter's identity, pretend guesses made sense given the limited information players pretended to have.

>>>>

---
title             : "Pretending not to know reveals a capacity for model-based self-simulation"
shorttitle        : "Pretending not to know"

author: 
  - name          : "Matan Mazor"
    affiliation   : "1"
    corresponding : yes    
    address       : "All Souls College, High Street, Oxford OX1 4AL"
    email         : "matan.mazor@all-souls.ox.ac.uk"

@@

  - name          : "Ian Phillips"
    affiliation   : "2"

@@

abstract: |
  Pretending not to know requires appreciating how one would behave without a given piece of knowledge, and acting accordingly. Here, two game-based experiments reveal a capacity to simulate decision-making under such counterfactual ignorance. 1001 English-speaking adults saw the solution to a game (ship locations in Battleship, the hidden word in Hangman) but attempted to play as though they never had this information. Pretenders accurately mimicked broad aspects of genuine play, including the number of guesses required to reach a solution, as well as subtle patterns, such as effects of decision uncertainty on decision time. While peers were unable to detect pretense, statistical analysis and computational modeling uncovered traces of ‘over-acting’ in pretenders’ decisions, suggesting a schematic simulation of their minds. Opening up a new approach to studying self-simulation, our results reveal intricate metacognitive knowledge about decision-making, drawn from a rich—but simplified—internal model of cognition.

keywords          : "pretense; metacognition; theory of mind"
wordcount         : "7745"

@@

classoption       : "man"
output            : papaja::apa6_word

bibliography: references.bib
---

```{r analysis-preferences, warning=F, message=F, echo=FALSE}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed, warning=F, message=F)

@@

# Load workspace (after running preregisteredMethodsAndResults.Rmd, 
# exploratoryResults.Rmd, and all the scripts that are linked to from these 
# documents). 
load('prereg_results.RData')
load('explor_results.RData')

papaja::r_refs("references.bib")

@@

# Research Transparency Statement

## General Disclosures

The authors declare no conflicts of interest. Funding: This study was supported by an NSF BCS grant #2021053 awarded to C.F. M.M is supported by a post-doctoral research fellowship at All Souls College. Artificial intelligence: No artificial intelligence assisted technologies were used in this research or the creation of this article. Ethics: The research complied with all relevant ethical regulations and was approved by the Institutional Review Board of Johns Hopkins University.

## Experiment 1 (Battleship)

Preregistration: The hypotheses and methods were preregistered on 2021-12–21, prior to data collection, and following the analysis of an independent pilot sample. A detailed pre-registration can be accessed at <https://osf.io/v9zsb>. The pre-registration was time-locked using cryptographic randomisation-based time-locking (Mazor, Mazor, & Mukamel, 2019) (protocol sum: 60c270410375e8a192468fc1a0e9c93da60d5e203eb2760b621a8631a26f4c5c; [link to relevant lines in experimental code](https://github.com/self-model/pretendingNotToKnow/blob/44ef964c0d7d4a8f3b76ec52316a8e624c41a047/experiments/Battleships2/code/index.html#L865-L876), making experimental randomisation causally dependent on the content of the pre-registration and thus ensuring that all data were collected after pre-registration. All pre-registered analyses are publicly available, including the report-generating R script (<https://self-model.github.io/pretendingNotToKnow/docs/exp.-1-battleship.html>). Exploratory analyses are flagged as such. Materials: All study materials, including demos of analysis experiments, are publicly available (<https://github.com/self-model/pretendingNotToKnow>). Data: All primary data are publicly available (<https://osf.io/zma9b>). Analysis scripts: All analysis scripts are publicly available (<https://osf.io/zma9b>).

## Experiment 2 (Hangman)

Preregistration: The hypotheses and methods were preregistered on 2022-06–22, prior to data collection, and following the analysis of an independent pilot sample. A detailed pre-registration can be accessed at osf.io/3thry. The pre-registration was time-locked using cryptographic randomisation-based time-locking, making experimental randomisation causally dependent on the content of the pre-registration and thus ensuring that all data were collected after pre-registration [@mazor2019]. Due to an error in the experiment code, time-locking took effect only from player number 221 (batch 3) and on (protocol sum: c4929c7fe33df1b7b52f15c789d98eab30a9cee09a8121807a3c59e28e7430a4;[relevant lines in experimental code](https://github.com/self-model/pretendingNotToKnow/blob/f98903f3f2d0e21800497dc65f8d5f823c910c88/experiments/Hangman2/code/index.html#L772-L783). All pre-registered analyses are publicly available, including the report-generating R script (<https://self-model.github.io/pretendingNotToKnow/docs/exp.-2-hangman.html>). Exploratory analyses are flagged as such. Materials: All study materials, including demos of the experiments, are publicly available (<https://github.com/self-model/pretendingNotToKnow>). Data: All primary data are publicly available (<https://osf.io/zma9b>). Analysis scripts: All analysis scripts are publicly available (<https://osf.io/zma9b>).

@@

Pretense relies on an ability to simulate and mimic one's own behavior under a counterfactual belief state. For example, in order to successfully deceive your friends into thinking that you were surprised by the birthday party they threw for you, it is not sufficient that you are able to reason about their mental states ("*I know that they are planning a surprise party, but they don't know that I know that...*") --- you also need to convincingly simulate and mimic your hypothetical behavior had you not known about the party ("*Where would I look first had I not known? What would I say? How long would it take me to recover from the surprise?*"). Similar examples abound in higher-stakes contexts such as diplomacy, warcraft and law. This is not a trivial challenge: previous research on "hindsight biases" suggests that knowledge about the actual state of the world can interfere with our ability to correctly judge what we would have believed [@fischhoff1975; @fischhoff1977; @wood1978; @roese2012] or perceived [@harley2004; @bernstein2007; @bernstein2012] without this knowledge. Such biases remain potent even when instructing participants to overcome them [@harley2004; @pohl1996]. Moreover, even if pretenders can correctly determine what they would have believed, they must further accurately simulate how they would think and behave in this different belief state.

@@

Our research question is whether people can mentally simulate their actions under a counterfactual knowledge state of ignorance. To that end, we had participants “pretend not to know” in a game setting. Using an online version of the games Battleship and Hangman (in which players seek to uncover the locations of enemy ships or the identity of a word), participants played a ‘non-pretend’ (normal) version of the game, as well as a ‘pretend’ version where they were given complete information about the hidden ships / target word but were instructed to behave as if they didn’t have this information. Participants’ pretense behaviour mirrored broad patterns and subtle features of real players’ decisions and decision times. At the same time, epistemic pretense was characterized by over-acting, stereotypical behavior, and suboptimal incorporation of new information: all markers of model-based simulation. Together, we take these findings as evidence for a capacity to mentally simulate decisions and actions using a simplified and schematic self-model.

# Method

The research complied with all relevant ethical regulations and was approved by the Institutional Review Board of Johns Hopkins University. In two experiments, online participants played online versions of two information-seeking games. In Battleship (Exp. 1), `r E2.df$subj_id%>%unique()%>%length()` English-speaking players (recruited from Prolific.com) were presented with a 5x5 grid of yellow squares, and attempted to reveal one size-3 submarine and two size-2 patrol boats with as few guesses as possible. In our version of the game, ships could only touch corner-to-corner, but not side-to-side (this was explained to participants before playing), and participants were not notified once they had sunk a ship (only whether their guess was a hit or a miss). In Hangman, N=`r E4.df$subj_id%>%unique()%>%length()` English-speaking players attempted to reveal a hidden word, name or number (hereafter referred to broadly as "word') with as few letter-guesses as possible, based on word length and category (a famous person, number, fruit, US state, or body part). To ensure familiarity with US states, Hangman participants were all US-based.

Both games traditionally start in a state of ignorance, with a player’s goal being to reveal an unknown world state (ship locations in Battleship, a hidden word in Hangman) in as few steps (cell or letter selections) as possible. Critically, in addition to playing five standard games, players in our experiments also completed five ‘pretend’ games in which the solution to the game was known to them from the start and remained visible on the screen throughout the entire game (pretend-Battleship ship locations were marked with a cross, pretend-Hangman words were presented visually and had to be typed by players before the game, to ensure encoding; see Fig. \@ref(fig:design)). In these games, players’ task was to behave as if they were playing for real – i.e. to play as though they did not have this information.

Throughout the game, participants accrued points that were later converted to a monetary bonus. In non-pretend games, participants received points for revealing the ships, or the target word, with as few guesses as possible. In pretend games, participants were given different instructions:

> “In this round, we’re going to tell you where the ships are, but we want you to act like you don’t know this information. We’ve marked the ships’ locations with a cross, so you’ll know where they are the whole time; but your job is to play the game as if these hints aren’t there. To see how good you are at this, we’re going to compare your games to the games of people who actually had no hints, and see how similar they are. We will measure where and when you clicked; if your clicks look similar to people who played like normal (trying to reveal all ships with as few clicks as possible, but without any hints), you’ll get bonus points. But if your games look different, you won’t get these bonus points. Your number of clicks in this part will not affect your bonus. Only your ability to play like you had no hints.”

And in Hangman:

> “In this round, we’re going to tell you the word in advance, but we want you to act like you don’t know this information. To see how good you are at this, we’re going to compare your games to the games of people who played normally, without knowing what the word was, and see how similar they are. We will measure which letters you click and the timing of your guesses; if your clicks look similar to people who played like normal (trying to reveal the word with as few guesses as possible, but without any hints), you’ll get bonus points. But if your games look different, you won’t get these bonus points. Your number of clicks in this part will not affect your bonus. Only your ability to play like you didn’t see the word in advance.”

We intentionally included no reference to an observer in these instructions, to have participants focusing on simulating their own behaviour rather than simulating how their behaviour would be perceived by another person. In reality, participants’ games were presented to other participants, and they received bonus points if they tricked these other participants into believing they did not have hints.

Players played pretend and standard games in separate blocks that were presented in random order after a first ‘practice’ game. In principle, participants could learn about their own behaviour from this practice game. To minimize such learning effects, we distinguished practice games from the main experimental blocks, using a smaller 4x4 grid with only two size-2 ships in Battleship, and a word category (animals) that was not used in the main experiment in Hangman. Each experimental block was followed by a half-game, where players were instructed to complete the game from a half-finished state. Finally, players were presented with replays of the games of previous players, and judged which were standard and which were pretend games. We measured players’ capacity to simulate a counterfactual state of ignorance by comparing patterns of decisions and decision times in pretend and non-pretend games. Our full pre-registered results are available [online](https://self-model.github.io/pretendingNotToKnow/docs/exp.-1-battleship.html) together with the report-generating code. Unless otherwise specified, all reported findings similarly hold when analysing only the first condition performed by each subject in a between-subject analysis, thereby ensuring that findings are not driven by learning effects^[In both experiments, the order of pretend and non-pretend blocks was counterbalanced between participants. We observed no significant interaction between the strength of any of our effects (i.e., differences between pretend and non-pretend conditions) and part of the experiment (i.e., first versus second). For full details see Supplementary Materials.]. Readers are invited to try [demos of the experiments](https://self-model.github.io/pretendingNotToKnow/experiments/demos/pretend).

@@

@@

As a first measure of pretense quality, we compared the total number of guesses in pretend and non-pretend games. Among Battleship players, the number of cell selections was similar in pretend (mean = `r E2.median_clicks%>%pull(pretend)%>%mean()%>%printnum()`, SD = `r E2.median_clicks%>%pull(pretend)%>%sd()%>%printnum()`) and non-pretend games (mean = `r E2.median_clicks%>%pull(nonpretend)%>%mean()%>%printnum()`, SD = `r E2.median_clicks%>%pull(nonpretend)%>%sd()%>%printnum()`; `r apa_print(E2.median_clicks%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$statistic`; Cohen's $d$ = `r E2.median_clicks%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%cohensD()%>%printnum()`; Fig. \@ref(fig:decisions)A). `r E2.lucky_pretenders%>%length` pretenders who immediately discovered all ships without making errors were excluded from all further analyses, in accordance with our pre-registered plan. With these subjects excluded, the number of cell selections remained very similar in pretend (mean = `r  E2.median_clicks_filtered%>%pull(pretend)%>%mean()%>%printnum()`) and non-pretend games (mean = `r E2.median_clicks_filtered%>%pull(nonpretend)%>%mean()%>%printnum()`; `r apa_print(E2.median_clicks_filtered%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$statistic`; Fig. \@ref(fig:decisions)A). In Hangman, pretenders tended to make about one additional letter guess on average than did non-pretenders, controlling for word length (pretend: `r E4.median_misses%>%pull(pretend)%>%mean()%>%printnum()` misses, SD = `r E4.median_misses%>%pull(pretend)%>%sd()%>%printnum()`; non-pretend: `r E4.median_misses%>%pull(nonpretend)%>%mean()%>%printnum()` misses, , SD = `r E4.median_misses%>%pull(nonpretend)%>%sd()%>%printnum()`; `r apa_print(E4.median_misses%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$statistic`; Cohen's $d$ = `r E4.median_misses%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%cohensD()%>%printnum()`; Fig. \@ref(fig:decisions)B). Despite an overall bias in the number of guesses, pretend Hangman games showed a near-perfect item-specific alignment: pretenders were successful in making more incorrect letter guesses when attempting to reveal words that would have been harder to guess had they been playing for real (`r cor.test(E4.misses_per_word$pretend, E4.misses_per_word$nonpretend,method='spearman')%>%apa_print()%>%'$'(estimate)`; Fig. \@ref(fig:decisions)B). This strong correlation provides evidence for a human capacity to act in accordance with a counterfactual knowledge state.

(ref:decisions-caption) Battleship and Hangman guesses in pretend and non-pretend games. A: median number of misses in Battleship and Hangman games, in non-pretend (green) and pretend (purple) games. For reference, the expected number of misses is indicated by a reference line for a fully random agent, and for a "greedy" agent that maximizes the probability of a hit in each step. B: The median number of misses in Hangman for pretend and non-pretend games, as a function of the target word. C: Spatial guess distributions for pretend and non-pretend half-games (where players continued the game from a half-finished state) alongside their corresponding hit probability maps. D: Cell and letter selections were ranked according to their relative hit probability given the players’ knowledge at the time of making the decision (dynamically updated after each guess). Plotting the median rank per subject in pretend and non-pretend games, with reference lines for the expected rank probability for a random agent, and for a “greedy” agent that maximizes the probability of a hit in each step. Note that the expected rank for a greedy agent is greater than 1 because there was not always a single optimal choice. E: same as panel D but discarding all guesses that resulted in a hit.

@@

num_misses_df <- E2.click_df %>%
  dplyr::select(subj_id,test_part, num_clicks) %>%
  mutate(subj_id=as.character(subj_id)) %>%
  mutate(test_part = factor(test_part, levels=c(
    'nonpretend','pretend'), labels = c(
    'non-pretend','pretend'))) %>%
  group_by(subj_id,test_part) %>%
  summarise(num_misses=median(num_clicks)-7) %>%
  mutate(game='Battleship') %>%
  rbind(
    E4.click_df %>%
  dplyr::select(subj_id,test_part, num_misses) %>%
  mutate(subj_id=as.character(subj_id)) %>%
  mutate(test_part = factor(test_part, levels=c(
    'nonpretend','pretend'), labels = c(
    'non-pretend','pretend'))) %>%
  group_by(subj_id,test_part) %>%
  summarise(num_misses=median(num_misses)) %>%
    mutate(game='Hangman')
  )
load('../analysis/simulated_hangman_games.RData')

@@

p<-  E2.click_df %>%
  dplyr::select(subj_id,test_part, num_clicks) %>%
  mutate(subj_id=as.character(subj_id)) %>%
  mutate(test_part = factor(test_part, levels=c(
    'nonpretend','pretend'), labels = c(
    'non-pretend','pretend'))) %>%
  group_by(subj_id,test_part) %>%
  summarise(num_misses=median(num_clicks)-7) %>%
  ggplot(aes(x=num_misses,fill=test_part, color=test_part))+
  scale_fill_manual(values=c(colors$np,colors$p, colors$random, colors$greedy)) +
  scale_x_continuous(limits=c(-1,19), breaks=seq(0,18)) +
  labs(x='median number of misses', y='number of players')+
  theme(legend.position='none') +
  theme_classic() +
  geom_vline(xintercept = E2.greedy_mean_misses, color='black', linetype=2) +
  annotate("text", x = E2.greedy_mean_misses, y = 75, label = "Greedy", angle = 90, vjust = -1, size = 5)+
  geom_vline(xintercept = E2.random_mean_misses, color='black', linetype=2) +
  annotate("text", x = E2.random_mean_misses, y = 75, label = "Random", angle = 90, vjust = -1, size = 5)+
  geom_bar(alpha=0.6, position = 'identity', color='black', size=0.5) +
  theme(legend.pos='na')

@@

Having established an alignment in the total number of guesses, we next turned to the content of pretend and non-pretend guesses. In order to directly compare pretend and non-pretend guesses for the same board state, Battleship players completed two half-games in which they were instructed to continue the game from a half-completed state. In standard games, players start in the same (blank) board state but quickly diverge as they make different guess sequences. Including half games allowed us access to hundreds of cell selections for the same board state from pretenders and non-pretenders. This way, we had sufficient statistical power to compare the two guess distributions. We find a strong correlation between the spatial distributions of pretend and non-pretend guesses (board A: $r=$ `r apa_print(cor_A_pretend_nonpretend)$table$estimate`, $p$ `r apa_print(cor_A_pretend_nonpretend)$table$p.value`; board B: $r=$ `r apa_print(cor_B_pretend_nonpretend)$table$estimate`, $p$ `r apa_print(cor_B_pretend_nonpretend)$table$p.value`; Fig. \@ref(fig:decisions)C), confirming that pretenders were sensitive not only to the number of guesses they would have made had they been playing for real, but also to their content.

To further examine the decisional processes behind this strong alignment, we compared the degree to which pretend and non-pretend guesses made sense within the context of the game. When playing Battleship and Hangman, it makes sense to guess cells or letters for which the probability of hitting a ship or revealing a letter is high [this "greedy" behavior is not strictly optimal, but approximates optimal behavior in most cases, @audinot2014optimal]. To this end, we ranked cells based on the Bayesian probability of a hit given players’ knowledge at the time of making the decision. Critically, hit probability maps were dynamically updated after each guess. In Battleship, this model assumed that all legal board configurations are equally likely a priori, but board configurations were ruled out as the game progressed and the content of individual cells was revealed. In Hangman, we used the category information (e.g., ‘a fruit’), to obtain a probability-weighted list of category-compatible words (or names, in the case of famous people). We relied on prototypicality norms [@uyeda1980prototypicality] for words, and number of visits to Wikipedia entries for famous people. The full prior distributions for each category were included in the pre-registration (for details, see Supplementary Materials). Similar to Battleship, in deriving hit probability we assume access to the full list of options that is consistent with the game state (the number of hidden letters, the revealed letters and their positions, and the list of letters that do not appear in the game solution) at the time of making the decision.
